---
title: "Project 1- Preprocessing"
authors: "Eva Martín, Laura González"
output: html_notebook
---

#Libraries
```{r}
library(dplyr)
library(missMDA)
library(FactoMineR)
library(car)
library(mice)
library (chemometrics)
library(MASS)
library(AER)
library(effects)
library(lmtest)
library(DescTools)
library(ResourceSelection)
library(statmod)
library(cvAUC)
library(caret)
library(ModelMetrics)
library(pROC)

```
#Load dataset and create dataframe
```{r}
df <- read.csv("C:/Users/lglez/Documents/MDS/Q3/AML/AML-Project1/data/bank-full.csv",
               sep = ";")

summary(df)

df_no_dupl <- distinct(df) #there are no duplicates in our dataset 
rm(df_no_dupl)


```

#Exploratory analisis
Assign correct datatype to each variable
No missing values, so no imputation method is required
```{r}
sapply(df, class)

# Convert characaters to categorical variables 
df[sapply(df, is.character)] <- lapply(df[sapply(df, is.character)], as.factor)
summary (df)

```
#Variables. From the dataset source:
1 - age (numeric)
487 outliers, all on the upper side. 3 of them severe. Not normally distributed
```{r}
summary(df[,1])

hist(df$age, breaks = 40, freq = F);curve(dnorm(x, mean(df$age), sd(df$age)), add = T)

Boxplot(df$age)
length(Boxplot(df$age, id=list(n=Inf)))

varsumm1<-summary (df[,1])
iqr1 <- varsumm1[5]-varsumm1[2]

umout1 <- varsumm1[5]+1.5*iqr1; umout1
usout1 <- varsumm1[5]+3*iqr1; usout1

lmout1 <- varsumm1[2]-1.5*iqr1; lmout1
lsout1 <- varsumm1[2]-3*iqr1; lsout1

mout1<- which((df[,1]<lmout1)|(df[,1]>umout1));length(mout1) #487 mout
sout1 <-which((df[,1]<lsout1)|(df[,1]>usout1));length(sout1) #3 sout

```

2 - job : type of job (categorical:"admin.","unknown","unemployed","management","housemaid","entrepreneur","student","blue-collar","self-employed","retired","technician","services")
12 levels
```{r}
summary(df[,2])
str(df$job)
plot(df$job) 

```


3 - marital : marital status (categorical: "married","divorced","single"; note: "divorced" means divorced or widowed)
```{r}
summary(df[,3])
str(df$marital)
plot(df$marital) 
```

4 - education (categorical: "unknown","secondary","primary","tertiary")
```{r}
summary(df[,4])
str(df$education)
plot(df$education)
```

5 - default: has credit in default? (binary: "yes","no")
```{r}
summary(df[,5])
str(df$default)
plot(df$default)
```
6 - balance: average yearly balance, in euros (numeric)
about 15% of outliers. canidate for future categorical variable
```{r}
summary(df[,6])

hist(df$balance, breaks = 40, freq = F);curve(dnorm(x, mean(df$balance), sd(df$balance)), add = T)

Boxplot(df$balance)
length(Boxplot(df$balance, id=list(n=Inf)))

varsumm6<-summary (df[,6])
iqr6 <- varsumm6[5]-varsumm6[2]

umout6 <- varsumm6[5]+1.5*iqr6; umout6
usout6 <- varsumm6[5]+3*iqr6; usout6

lmout6 <- varsumm6[2]-1.5*iqr6; lmout6
lsout6 <- varsumm6[2]-3*iqr6; lsout6

mout6<- which((df[,6]<lmout6)|(df[,6]>umout6));length(mout6) #4729 mout
sout6 <-which((df[,6]<lsout6)|(df[,6]>usout6));length(sout6) #2443 sout
```

7 - housing: has housing loan? (binary: "yes","no")
```{r}
summary(df[,7])
str(df$housing)
plot(df$housing)
```


8 - loan: has personal loan? (binary: "yes","no")
```{r}
summary(df[,8])
str(df$loan)
plot(df$loan)
```

# related with the last contact of the current campaign:
9 - contact: contact communication type (categorical: "unknown","telephone","cellular")
```{r}
summary(df[,9])
str(df$contact)
plot(df$contact)
```

10 - day: last contact day of the month (numeric)
doesn't make sense to treat it as a numeric value
```{r}
df$day <- as.factor(df$day)
summary(df[,10])
str(df$day)
plot(df$day)

```

11 - month: last contact month of year (categorical: "jan", "feb", "mar", …, "nov", "dec")
circular variable (after dec -> jan again)
```{r}
# Order
df$month <- factor(df$month, 
                   levels = c("jan","feb","mar","apr","may","jun",
                              "jul","aug","sep","oct","nov","dec"),
                   ordered = TRUE)

#Cyclic codification
df$month_sin <- sin(2*pi*as.numeric(df$month)/12)
df$month_cos <- cos(2*pi*as.numeric(df$month)/12)

summary(df[,11])    
str(df$month)


plot(df$month)

```

12 - duration: last contact duration, in seconds (numeric)
not normally dist
```{r}
summary(df[,12])

hist(df$duration, breaks = 40, freq = F);curve(dnorm(x, mean(df$duration), sd(df$duration)), add = T)

Boxplot(df$duration)
length(Boxplot(df$duration, id=list(n=Inf)))

varsumm12<-summary (df[,12])
iqr12 <- varsumm12[5]-varsumm12[2]

umout12 <- varsumm12[5]+1.5*iqr12; umout12
usout12<- varsumm12[5]+3*iqr12; usout12

lmout12<- varsumm12[2]-1.5*iqr12; lmout12
lsout12<- varsumm12[2]-3*iqr12; lsout12

mout12<- which((df[,12]<lmout12)|(df[,12]>umout12));length(mout12) #3235 mout
sout12<-which((df[,12]<lsout12)|(df[,12]>usout12));length(sout12) #1155 sout

```

# other attributes:
13 - campaign: number of contacts performed during this campaign and for this client (numeric, includes last contact)
```{r}
summary(df[,13])

hist(df$campaign, breaks = 40, freq = F);curve(dnorm(x, mean(df$campaign), sd(df$campaign)), add = T)

Boxplot(df$campaign)
length(Boxplot(df$campaign, id=list(n=Inf)))

varsumm13<-summary (df[,13])
iqr13 <- varsumm13[5]-varsumm13[2]

umout13 <- varsumm13[5]+1.5*iqr13; umout13
usout13<- varsumm13[5]+3*iqr13; usout13

lmout13<- varsumm13[2]-1.5*iqr13; lmout13
lsout13<- varsumm13[2]-3*iqr13; lsout13

mout13<- which((df[,13]<lmout13)|(df[,13]>umout13));length(mout13) #3064 mout
sout13<-which((df[,13]<lsout13)|(df[,13]>usout13));length(sout13) #1462 sout

```

14 - pdays: number of days that passed by after the client was last contacted from a previous campaign (numeric, -1 means client was not previously contacted)
problematic variable. addressed in 2 different variables:
- binary value if was contacted or not 
- replace -1 with nas: 82 % of NAs -> remove feature
```{r}
df$was_contacted <- ifelse(df$pdays == -1, 0, 1)
df$was_contacted <- as.factor(df$was_contacted)
summary(df$was_contacted)
str(df$was_contacted)
plot(df$was_contacted)

df$pdays <- ifelse(df$pdays == -1, NA, df$pdays)
summary (df$pdays)
df <- subset(df, select = -pdays)
summary(df)

```


15 - previous: number of contacts performed before this campaign and for this client (numeric)
```{r}
summary(df[,14])

hist(df$previous, breaks = 40, freq = F);curve(dnorm(x, mean(df$previous), sd(df$previous)), add = T)

Boxplot(df$previous)
length(Boxplot(df$previous, id=list(n=Inf)))

varsumm14<-summary (df[,14])
iqr14 <- varsumm14[5]-varsumm14[2]

umout14 <- varsumm14[5]+1.5*iqr14; umout14
usout14<- varsumm14[5]+3*iqr14; usout14

lmout14<- varsumm14[2]-1.5*iqr14; lmout14
lsout14<- varsumm14[2]-3*iqr14; lsout14

mout14<- which((df[,14]<lmout14)|(df[,14]>umout14));length(mout14) #8257 mout
sout14<-which((df[,14]<lsout14)|(df[,14]>usout14));length(sout14) #8257 sout
```


16 - poutcome: outcome of the previous marketing campaign (categorical: "unknown","other","failure","success")
```{r}
summary(df[,15])
str(df$poutcome)
plot(df$poutcome)
```


Output variable (desired target):
17 - y - has the client subscribed a term deposit? (binary: "yes","no")

```{r}
summary(df[,16])
str(df$y)
plot(df$y)

```

